---
title: "MB5370-Marine data analysis- Coral Bleaching"
output: html_document
date: "2026-06-28"
---

## Introduction
This project will be about the bleaching severity along different coral species throughout regions of the world. In hoping to assess how parts of the world have different thriving coral species and communities. It would be interesting to observe similarities in how affected coral species are during bleaching events per year. 


```{r}
coral_bleaching<-read.csv("C:/Users/jd131798/Downloads/Marine_data_analysis/Data/Coral_Bleaching.csv")
head(coral_bleaching)
```

```{r warning=FALSE}
install.packages("tidyverse")
library(tidyverse)
library(dplyr)
library(ggplot2)
```
```{r}
coral_bleaching_clean <-  coral_bleaching %>%
  select(-"subregion", -"country", -"location", -"lat", -"lon", -"depth", -"water_temperature", -"country_code", -"source", -"water_temperature" , -"percentage_affected", -"coral_family", -"coral_species", -"mortality", -"mortality_code")
```



```{r}
coral_clean<- na.omit(coral_bleaching_clean)

```
```{r}
coral_clean <- coral_clean %>%
  mutate(bleaching_severity = na_if(bleaching_severity, "Unknown"))

coral_clean2<- na.omit(coral_clean_U)
```

```{r}
distinct(coral_clean2, bleaching_severity)
```


```{r}
coral_clean2|>
  ggplot(aes(x=year, y=severity_code, color=region))+
  geom_point(data=coral_clean2, aes(x=year, y=severity_code, color=region), 
             position=position_jitter(width=0.5, height=0.2), alpha=0.6, size=2, inherit.aes=FALSE)+
  geom_point(position="dodge", size=2, pch=20)+

  facet_wrap(~region, scales= "fixed")+
  ggtitle("Centered Plot Title") +
  theme_minimal()+
  theme(
    strip.text = element_text(size=12, face="bold", hjust = 0.5),
    legend.position="bottom"
  )+
  labs(
    title="Coral Bleaching Severity by Region and Year",
    x="Year",
    y="Severity Code",
    color="Region")
    
```
# this graph above displays the bleaching severity across different regions of the world. The points represent individual observations, with some jitter added to avoid overlap. The error bars indicate the range of severity for each year within each region. The facet wrap allows for a clear comparison of bleaching severity trends across different regions over time.
```{r}
Coral_bleaching_clean|>
  ggplot(aes(x=year, y=percentage_affected, color=region))+
  geom_point(data=Coral_bleaching_clean, aes(x=year, y=mortality, color=region), 
             position=position_jitter(width=0.5, height=0.2), alpha=0.6, size=2, inherit.aes=FALSE)+
  geom_point(position="dodge", size=2, pch=21)+
  
  geom_errorbar(aes(ymin=mortality, ymax=mortality), width=0.5, position=position_dodge(0.9))+
  facet_wrap(~region, scales= "fixed")+
  theme_minimal()+
  theme(
    strip.text = element_text(size=12, face="bold"),
    legend.position="bottom"
  )+
  labs(
    title="Coral mortality by Region and Year",
    x="Year",
    y="Mortality",
    color="Region")
  )
```
# this plot displays the mortality rates of coral across different regions over time. Similar to the previous plot, individual observations are represented by points with jitter to avoid overlap, and error bars indicate the range of mortality for each year within each region. The facet wrap allows for a clear comparison of mortality trends across different regions over time.
```{r}
select_coral_bleaching|>
  ggplot(aes(x=year, y=severity_code, color=region))+
  geom_point(data=select_coral_bleaching, aes(x=year, y=bleaching_severity, color=region), 
             position=position_jitter(width=0.5, height=0.2), alpha=0.6, size=2, inherit.aes=FALSE)+
  geom_point(position="dodge", size=2, pch=21)+
  scale_y_discrete(limits = c("Low", "Medium", "High")) +
  
  geom_errorbar(aes(ymin=bleaching_severity, ymax=bleaching_severity), width=0.5, position=position_dodge(0.9))+
  facet_wrap(~region, scales= "fixed")+
  ggtitle("Centered Plot Title") +
  theme_minimal()+
  theme(
    strip.text = element_text(size=12, face="bold", hjust = 0.5),
    legend.position="bottom"
  )+
  labs(
    title="Coral Bleaching Severity by Region and Year",
    x="Year",
    y="Severity Code",
    color="Region")
```

```{r}

```








