---
title: "MB5370-Marine data analysis- Coral Bleaching"
output: html_document
date: "2026-06-28"
---

## Introduction
This project will be about the bleaching severity along different coral species throughout regions of the world. In hoping to assess how parts of the world have different thriving coral species and communities. It would be interesting to observe similarities in how affected coral species are during bleaching events per year. 


```{r}
coral_bleaching<-read.csv("C:/Users/jd131798/Downloads/Marine_data_analysis/Data/Coral_Bleaching.csv")
head(coral_bleaching)
```

```{r warning=FALSE}
install.packages("tidyverse")
library(tidyverse)
library(dplyr)
library(ggplot2)
```

```{r}
coral_bleaching_clean <-  coral_bleaching %>%
  select(-"subregion", -"country", -"location", -"lat", -"lon", -"depth", -"water_temperature", -"country_code", -"source", -"water_temperature" , -"percentage_affected", -"coral_family", -"coral_species", -"mortality", -"mortality_code")
```


```{r}
coral_clean<- na.omit(coral_bleaching_clean)
```

```{r}
coral_clean <- coral_clean %>%
  mutate(bleaching_severity = na_if(bleaching_severity, "Unknown"))

coral_clean2<- na.omit(coral_clean)
```

```{r}
distinct(coral_clean2, bleaching_severity)
```


```{r}
coral_clean2|>
  ggplot(aes(x=year, y=severity_code, color=region))+
  geom_point(data=coral_clean2, aes(x=year, y=severity_code, color=region), 
             position=position_jitter(width=0.5, height=0.2), alpha=0.6, size=2, inherit.aes=FALSE)+
  geom_point(position="dodge", size=2, pch=20)+

  facet_wrap(~region, scales= "fixed")+
  ggtitle("Centered Plot Title") +
  theme_minimal()+
  theme(
    strip.text = element_text(size=12, face="bold", hjust = 0.5),
    legend.position="bottom"
  )+
  labs(
    title="Coral Bleaching Severity by Region and Year",
    x="Year",
    y="Severity Code",
    color="Region")
    
```
# this graph above displays the bleaching severity across different regions of the world. The points represent individual observations, with some jitter added to avoid overlap. The error bars indicate the range of severity for each year within each region. The facet wrap allows for a clear comparison of bleaching severity trends across different regions over time.


```{r}
#NO DON'T USE
coral_clean2|>
  ggplot(aes(x=year, y=severity_code, colour=region))+
  geom_violin(data=coral_clean2, aes(x=year, y=severity_code, fill=region), 
             position=position_jitter(width=0.5, height=0.2), alpha=0.6, size=2, inherit.aes=FALSE)+
  geom_point(position="dodge", size=1, pch=20)+

  facet_wrap(~region, scales= "fixed")+
  ggtitle("Centered Plot Title") +
  theme_minimal()+
  theme(
    strip.text = element_text(size=12, face="bold", hjust = 0.5),
    legend.position="bottom"
  )+
  labs(
    title="Coral Bleaching Severity by Region and Year",
    x="Year",
    y="Severity Code",
    color="Region")
```
```{r}
#NO DO NOT USE
coral_clean2|>
  ggplot(aes(x=year, y=severity_code, color=region))+
  geom_line(data=coral_clean2, aes(x=year, y=severity_code, color=region), 
             position=position_jitter(width=0.9, height=0.3), alpha=0.4, size=2, inherit.aes=FALSE)+
  geom_point(position="dodge", size=2, pch=20)+
  geom_line(linewidth = 1.0) +

  facet_wrap(~region, scales= "fixed")+
  ggtitle("Centered Plot Title") +
  theme_minimal()+
  theme(
    strip.text = element_text(size=12, face="bold", hjust = 0.6),
    legend.position="bottom"
  )+
  labs(
    title="Coral Bleaching Severity by Region and Year",
    x="Year",
    y="Severity Code",
    color="Region")
```
```{r}
coral_clean2|>
  ggplot(aes(x=year, y=severity_code, fill = region)) +
geom_violin(alpha = 0.5) +
geom_jitter(position = position_jitter(width = 0.2), alpha = 0.6) +
theme_minimal() +


  facet_wrap(~region, scales= "fixed")+
  ggtitle("Centered Coral Bleaching Severity by Region and Year") +
  theme_minimal()+
  theme(
    strip.text = element_text(size=11, face="bold", hjust = 0.3),
    legend.position="bottom"
  )+
  labs(
    title="Coral Bleaching Severity by Region and Year",
    x="Year",
    y="Severity Code",
    color="Region")
```

```{r}
ggplot(data = coral_clean2, aes(x=year, y=severity_code, fill = region)) +
  geom_boxplot()+

 facet_wrap(~region, scales= "fixed")+
  ggtitle("Centered Coral Bleaching Severity by Region and Year") +
  theme_minimal()+
  theme(
    strip.text = element_text(size=11, face="bold", hjust = 0.3),
    legend.position="bottom"
  )+
  labs(
    title="Coral Bleaching Severity by Region and Year",
    x="Year",
    y="Severity Code",
    color="Region")
```

```{r}
#listing the years in order to prepare to calculate the mean of the bleaching severity per region over time
year_order<-distinct(coral_clean2, year)
```

```{r}
aggregate(coral_clean2 ~ region, data = coral_clean2, FUN = mean)
```
```{r}
average_severity_by_region <- coral_clean2 %>%
  group_by(region, year) %>%
  summarise(mean_severity = mean(severity_code, na.rm = TRUE))
```

grouped_penguins <- group_by(penguins, species)

# Notice that the table looks identical, but metadata notes 'Groups: species [3]'
print(grouped_penguins)

# Collapsing the buckets into explicit summary metrics
species_mass_summary <- summarise(grouped_penguins,
  mean_mass_g = mean(body_mass_g)
)

print(species_mass_summary)

```{r}
average_severity_by_region|>
  ggplot(aes(x=year, y=mean_severity,  colour=region, size = mean_severity))+
  geom_point(data=average_severity_by_region, aes(x=year, y=mean_severity, colour=region), 
             position=position_jitter(width=0.5, height=0.2), alpha=0.7)+
  geom_point(position="dodge", size= 3.5, pch=20)+
  facet_wrap(~region)+

    scale_y_continuous(
    limits = c(0, 3),
    breaks = 0:3,
    labels = c("No Bleaching", "Low", "Medium", "High")
  ) +
  
  theme_minimal()+
  theme(
    strip.text = element_text(size=12, face="bold", hjust = 0.5),
    legend.position="bottom"
  )+
  labs(
    title="Coral Bleaching Severity by Region and Year",
    x="Year",
    y="Mean Bleaching Severity ",
    color="Regions")
```
```{r}
summary_table <- coral_clean2 %>%
  group_by(region, year) %>%
  summarise(
    mean_severity = mean(severity_code, na.rm = TRUE),
    n = n(),
    no_bleaching = sum(severity_code == 0),
    low = sum(severity_code == 1),
    medium = sum(severity_code == 2),
    high = sum(severity_code == 3),
    .groups = "drop"
  )

summary_table

```

```{r}
ggplot(summary_table,
       aes(x = year,
           y = mean_severity,
           colour = region,
           group = region)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~region) +
  scale_y_continuous(
    limits = c(0, 3),
    breaks = 0:3,
    labels = c("No Bleaching", "Low", "Medium", "High")
  ) +
  theme_bw() +
  labs(
    title = " Coral Bleaching Severity by Region and Year",
    x = "Year",
    y = " Bleaching Severity",
    colour = "Region"
  )
```

```{r}
#Reference code if needed
Coral_bleaching_clean|>
  ggplot(aes(x=year, y=percentage_affected, color=region))+
  geom_point(data=Coral_bleaching_clean, aes(x=year, y=mortality, color=region), 
             position=position_jitter(width=0.5, height=0.2), alpha=0.6, size=2, inherit.aes=FALSE)+
  geom_point(position="dodge", size=2, pch=21)+
  
  geom_errorbar(aes(ymin=mortality, ymax=mortality), width=0.5, position=position_dodge(0.9))+
  facet_wrap(~region, scales= "fixed")+
  theme_minimal()+
  theme(
    strip.text = element_text(size=12, face="bold"),
    legend.position="bottom"
  )+
  labs(
    title="Coral mortality by Region and Year",
    x="Year",
    y="Mortality",
    color="Region")
  )
```

```{r}
select_coral_bleaching|>
  ggplot(aes(x=year, y=severity_code, color=region))+
  geom_point(data=select_coral_bleaching, aes(x=year, y=bleaching_severity, color=region), 
             position=position_jitter(width=0.5, height=0.2), alpha=0.6, size=2, inherit.aes=FALSE)+
  geom_point(position="dodge", size=2, pch=21)+
  scale_y_discrete(limits = c("Low", "Medium", "High")) +
  
  geom_errorbar(aes(ymin=bleaching_severity, ymax=bleaching_severity), width=0.5, position=position_dodge(0.9))+
  facet_wrap(~region, scales= "fixed")+
  ggtitle("Centered Plot Title") +
  theme_minimal()+
  theme(
    strip.text = element_text(size=12, face="bold", hjust = 0.5),
    legend.position="bottom"
  )+
  labs(
    title="Coral Bleaching Severity by Region and Year",
    x="Year",
    y="Severity Code",
    color="Region")
```

```{r}

```








